---
title: "American Family Field Mapping"
author: "Paulina Leperi"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tigris)
library(sf)
library(crsuggest)
library(mapboxapi)
library(tidyverse)
library(tidycensus)
library(spdep)
options(tigris_use_cache = TRUE)
theme_set(theme_bw())
```

To begin, I'll acquire CRS data for American Family Field. In this dataset,
it's still called Miller Park.

```{r}
stadium_url <- "https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/MajorSportVenues/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"
stadium <- st_read(stadium_url) %>%
  filter(str_detect(str_to_upper(NAME), "MILLER PARK"))
```

Once the data is loaded, a recommendation is needed for the best CRS to use. 
All layers of the maps created in subsequent steps will be transformed
to the same CRS.

```{r}
suggest_crs(stadium)
```

I'll use 8159, which measures distance in meters.

```{r}
stadium_t <- st_transform(stadium, 8159)
```

Next, I want to look at the census tracts surrounding the stadium. I'll acquire 
them using Tidycensus and transform them to the same CRS, then filter to only
include the tracts within 50 km of the stadium. (Please note that use of 
Tidycensus requires an API key.)

```{r}
wi_tracts <- tracts("WI", cb = TRUE, year = 2021) %>%
  st_transform(8159) %>%
  st_filter(stadium_t, .predicate = st_is_within_distance, dist = 50000) %>%
  erase_water()
```

The final layer in my first visualization will show the area within a 30-minute
drive of the stadium. 

```{r}
iso30min <- mb_isochrone(
  stadium_t, 
  time = 30, 
  profile = "driving-traffic"
) %>%
  st_transform(8159)
```

Here's a map that layers on the three components generated so far.

```{r}
ggplot() + 
  geom_sf(data = wi_tracts, color = "grey50", fill = "lightgray") + 
  geom_sf(data = iso30min, color = "black", fill = "pink") +
  geom_sf(data = stadium_t, color = "red") + 
  theme_void()
```

The next thing I want to look at is the age of the population near the stadium.
The following code gets median age by census tract for the state of Wisconsin
and then filters the dataset to include only those tracts that intersect
the 30-minute drive polygon.

```{r}
age_by_tract <- get_acs(
  geography = "tract",
  variables = "B01002_001",
  state = "WI",
  year = 2021,
  geometry = TRUE
) %>%
  st_transform(8159) %>%
  st_filter(iso30min, .predicate = st_intersects) %>%
  na.omit()

ggplot(age_by_tract) + 
  geom_sf(aes(fill = estimate), color = NA) + 
  scale_fill_viridis_c() + 
  geom_sf(data = stadium_t, color = "red") + 
  theme_void() +
  labs(fill = "Median Age")
```

There appear to be areas to the northeast of the stadium where the median age
of the population is low, but this can be quantified in a more precise way.

I will use local spatial autocorrelation analysis to identify "hotspots"
where the population differs from total values in the dataset.

Analysis like this can help precisely target advertising for stadium
events and promotions.

```{r}
neighbors <- poly2nb(age_by_tract, queen = TRUE)
weights <- nb2listw(include.self(neighbors))
age_by_tract$localG <- localG(age_by_tract$estimate, weights)

age_by_tract <- age_by_tract %>%
  mutate(hotspot = case_when(
    localG >= 2.576 ~ "High cluster",
    localG <= -2.576 ~ "Low cluster",
    TRUE ~ "Not significant"
  ))

ggplot(age_by_tract) + 
  geom_sf(aes(fill = hotspot), color = "grey90", size = 0.1) + 
  scale_fill_manual(values = c("red", "cyan", "grey")) + 
  geom_sf(data = stadium_t, color = "black", size = 5) + 
  theme_void() +
  labs(title = "High/Low Median Age \nNear American Family Field",
       fill = "")

```